Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-28 manuscript.tex
latest

Stable coexistence is possible between genotypes that are superior in different traits. Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence, and are not present in the classic lottery model. Mutual invasibility is not possible between $b$- and $d$-specialists.

@@

Summing Eq. \eqref{eq:master} over genotypes, $N$ obeys 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
where $U=\sum_i b_i n_i/L$. In equilibrium ($\Delta N=0$), Eq. \eqref{eq:deltN} implies $L/(1-e^{-L})=\sum_i b_i n_i/\sum_i d_i n_i$, and so $L$ only depends on the ratio of population mean birth rate to population mean death rate.

We consider the invasion of mutants into a population with a single type $i$ in equilibrium. The fitness of low frequency invader types $j$ is
\begin{equation}
\Delta n_j/n_j = b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
\end{equation}
where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.

In ideal environments, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, \label{eq:invad}
\end{equation}
where without loss of generality we set the reference competitiveness at $c_i=1$.

%From Eq. \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$. Clearly $b_i$ must be exceptionally large to ensure population persistence. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

>>>>

Now suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), with $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
\begin{equation}
b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
\end{equation}
Thus, coexistence occurs if $c_j/c_i$ is large enough. This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence, and are not present in the classic lottery model. Mutual invasion is not possible between $b$- and $d$-specialists.

@@

In this Appendix we consider the invasion of mutants into a population with a single type $i$ in equilibrium under the different environments from section ``Primary strategies and Grime's triangle''. We first evaluate the ideal and stressful environments (the treatment of mortality is different for disturbed environments).

Summing Eq. \eqref{eq:master} over genotypes, the change in $N$ is given by  
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
where $U=\sum_i b_i n_i/L$. In equilibrium ($\Delta N=0$), this implies $L/(1-e^{-L})=b_i/d_i $. The fitness of a low frequency invader $j$ is
\begin{equation}
\Delta n_j/n_j \approx b_j \left(e^{-L}+R_jc_j\right)-d_j \label{eq:invad}
\end{equation}
where $R_j\approx (1-e^{-L})/\left(c_j+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$, and we set the reference competitiveness at $c_i=1$ without loss of generality.

According our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and thus $L\ll 1$. Then Eq. \eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.

According to our definition of an ideal environment, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, 
\end{equation}

From Eq. \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$. Clearly $b_i$ must be exceptionally large to ensure population persistence. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
